---
title: "final project"
author: "Karl Veggerby"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# load packages and data loading and organizing
```{r load packages and data loading and organizing, message=FALSE}
# required packages
library(tidyverse)
library(lme4)
library(MuMIn)
library(RColorBrewer)
library(here)
library(arm)
library(performance)

# load data
feed_behav<-"behavior_data.csv"
feed_behav <- here("data/clean_data", feed_behav)
feed_behav <- read_csv(feed_behav,
                na = c("", "NA"))


# change feeding behavior to be a factor
feed_behav$feeding_behavior<-as.factor(feed_behav$feeding_behavior)

# create a summary table to view species counts
feed_behav_sum<-feed_behav %>%
    group_by(Subject) %>%
  summarize(counts = length(Subject))

feed_behav_sum_species<-feed_behav %>%
    group_by(Subject,feeding_behavior) %>%
  summarize(feeding_count = length(Subject))

feed_behav_sum_groups<-feed_behav %>%
    group_by(vertical_distribution,feeding_behavior) %>%
  summarize(feeding_count = length(Subject))

sum_species<-feed_behav_sum_species %>%
    group_by(Subject)  %>%
  summarize(total_count = sum(feeding_count))


# group by video as well as all other variables interest to account for the effect of video segment
feed_behav<-feed_behav %>%
  group_by(ObservationID,
           Subject,
           habitat_type,
           Site,
           visibility,
           vertical_distribution,
           feeding_behavior,
           eelgrass_algae_bare)%>%
  summarize(counts = length(Subject))

# subset into vertical functional groups for analysis
pelagic<-filter(.data =feed_behav,vertical_distribution == "pelagic/free swimming")
  
demersal<-filter(.data =feed_behav,vertical_distribution == "demersal")
  
benthic<-filter(.data =feed_behav,vertical_distribution == "benthic")  

```



# Pelagic organisms model
```{r pelagic model selection}

# determine random error structure
pelagic_model_no_random <- glm(feeding_behavior ~ habitat_type + eelgrass_algae_bare + visibility, data = pelagic, family = binomial)


pelagic_model_with_random <- glmer(feeding_behavior ~ habitat_type + eelgrass_algae_bare + visibility + (1 | Site), data = pelagic, family = binomial)


model.sel(pelagic_model_no_random,pelagic_model_with_random)
# random effect of site improves model performance. Will proceed with site in all models.

# perform full model selection 
options(na.action = "na.fail") 

full_model_pelagic <- glmer(feeding_behavior ~ habitat_type + eelgrass_algae_bare + visibility + (1 | Site), data = pelagic, family = binomial)



ms1 <- dredge(global.model = full_model_pelagic, evaluate = TRUE,fixed = c("habitat_type","Site"))

# top pelagic model
top_model_pelagic<-full_model_pelagic <- glmer(feeding_behavior ~ habitat_type + eelgrass_algae_bare + (1 | Site), data = pelagic, family = binomial)

```

pelagic diagnostics
```{r pelagic diagnostics}
## plot resids vs fitted
plot(fitted(top_model_pelagic), residuals(top_model_pelagic), pch = 16,
     xlab = "Fitted", ylab = "Residuals",
     main = "Residuals vs fitted")
abline(h=0, lty = "dashed")


# QQplot look for unusual points
qqplot(fitted(top_model_pelagic), residuals(top_model_pelagic))


# Deviance residuals: Plot the binned fitted versus residuals
binnedplot(fitted(top_model_pelagic), residuals(top_model_pelagic), las = 1, pch = 16,
     ylab = "Residuals", xlab = "Fitted values",
     main = "")


# r^2
r2_nakagawa(top_model_pelagic)


# overdispersion test
nn<-392

## overdispersion parameter
c_hat <- deviance(top_model_pelagic) / (nn - 2)
## re-scaled estimates

## 1.147118
# dispersion parameter close to 1


## goodness of fit test

## Pearson's X^2 statistic
nn<-392
X2 <- sum(((as.numeric(pelagic$feeding_behavior)-1) - fitted(top_model_pelagic))^2 / fitted(top_model_pelagic))
## likelihood ratio test
pchisq(X2, df = nn - length(coef(top_model_pelagic)),
lower.tail = FALSE)

# 1
# null Ho is that the model is correctly specified 
# p value is large so we detect no lack of fit
```



# Demersal organisms model
```{r demersal model selection}

# determine random error structure
demersal_model_no_random <- glm(feeding_behavior ~ habitat_type + eelgrass_algae_bare + visibility, data = demersal, family = binomial)


demersal_model_with_random <- glmer(feeding_behavior ~ habitat_type + eelgrass_algae_bare + visibility + (1 | Site), data = demersal, family = binomial)


model.sel(demersal_model_no_random,demersal_model_with_random)
# random effect of site decreases model performance. Will proceed without site in all models.


# perform full model selection 
options(na.action = "na.fail") 

full_model_demersal <- glm(feeding_behavior ~ habitat_type + eelgrass_algae_bare + visibility, data = demersal, family = binomial)



ms2 <- dredge(global.model = full_model_demersal, evaluate = TRUE,fixed = "habitat_type")

# top demersal model
top_model_demersal <- glm(feeding_behavior ~ habitat_type, data = demersal, family = binomial)

```

demersal diagnostics
```{r demersal diagnostics}
## plot resids vs fitted
plot(fitted(top_model_demersal), residuals(top_model_demersal), pch = 16,
     xlab = "Fitted", ylab = "Residuals",
     main = "Residuals vs fitted")
abline(h=0, lty = "dashed")


# QQplot look for unusual points
qqplot(fitted(top_model_demersal), residuals(top_model_demersal))


# Deviance residuals: Plot the binned fitted versus residuals
binnedplot(fitted(top_model_demersal), residuals(top_model_demersal), las = 1, pch = 16,
     ylab = "Residuals", xlab = "Fitted values",
     main = "")


# calcuate R^2
r2_coxsnell(top_model_demersal)


# overdispersion test

nn<-269
## overdispersion parameter
c_hat <- deviance(top_model_demersal) / (nn - 2)
## re-scaled estimates
faraway::sumary(top_model_demersal, dispersion = c_hat)
## 0.95085

# dispersion parameter less than 1


## Pearson's X^2 statistic
nn<-269
X2 <- sum(((as.numeric(demersal$feeding_behavior)-1) - fitted(top_model_demersal))^2 / fitted(top_model_demersal))
## likelihood ratio test
pchisq(X2, df = nn - length(coef(top_model_demersal)),
lower.tail = FALSE)

# 0.9918447
# null Ho is that the model is correctly specified 
# p value is large so we detect no lack of fit
```



# benthic organisms model
```{r benthic model selection}

# determine random error structure
benthic_model_no_random <- glm(feeding_behavior ~ habitat_type + eelgrass_algae_bare + visibility, data = benthic, family = binomial)


benthic_model_with_random <- glmer(feeding_behavior ~ habitat_type + eelgrass_algae_bare + visibility + (1 | Site), data = benthic, family = binomial)


model.sel(benthic_model_no_random,benthic_model_with_random)
# random effect of site decreases model performance. Will proceed without site in all models.


# perform full model selection 
options(na.action = "na.fail") 

full_model_benthic <- glm(feeding_behavior ~ habitat_type + eelgrass_algae_bare + visibility, data = benthic, family = binomial)



ms3 <- dredge(global.model = full_model_benthic, evaluate = TRUE,fixed = "habitat_type")

# top benthic model
top_model_benthic <- glm(feeding_behavior ~ habitat_type + visibility + eelgrass_algae_bare, data = benthic, family = binomial)

```

benthic diagnostics
```{r benthic diagnostics}
## plot resids vs fitted
plot(fitted(top_model_benthic), residuals(top_model_benthic), pch = 16,
     xlab = "Fitted", ylab = "Residuals",
     main = "Residuals vs fitted")
abline(h=0, lty = "dashed")


# QQplot look for unusual points
qqplot(fitted(top_model_benthic), residuals(top_model_benthic))


# Deviance residuals: Plot the binned fitted versus residuals
binnedplot(fitted(top_model_benthic), residuals(top_model_benthic), las = 1, pch = 16,
     ylab = "Residuals", xlab = "Fitted values",
     main = "")


# calcuate R^2
r2_coxsnell(top_model_benthic)


# overdispersion test

nn<-184
## overdispersion parameter
c_hat <- deviance(top_model_benthic) / (nn - 2)
## re-scaled estimates
faraway::sumary(top_model_benthic, dispersion = c_hat)
## 0.7407814

# dispersion parameter less than 1


## Pearson's X^2 statistic
nn<-184
X2 <- sum(((as.numeric(benthic$feeding_behavior)-1) - fitted(top_model_benthic))^2 / fitted(top_model_benthic))
## likelihood ratio test
pchisq(X2, df = nn - length(coef(top_model_benthic)),
lower.tail = FALSE)

# 0.5764506
# null Ho is that the model is correctly specified 
# p value is large so we detect no lack of fit
```





# plots of behavior across different habitats

model coefficients plots benthic
```{r model coefficients plots benthic}

benthic_model_top <- glm(feeding_behavior ~ 0 + habitat_type + visibility + eelgrass_algae_bare, data = benthic, family = binomial)

summary(benthic_model_top)$coefficients

# benthic

#                                       
#                                    Estimate  Std. Error
# habitat_typeeelgrass              -2.2085646    1.418819
# habitat_typesediment               0.3460405    1.143079
# habitat_typeFlipbag               -1.7925503    1.247346
# habitat_typeClam                  -0.5452848    1.294661 
# habitat_typeOyster OnBottom       -1.6230180    1.371282 
# 
# eelgrass_algae_bareeelgrass        2.3462651    1.1650095 
# eelgrass_algae_barealgae          -0.5452848    1.2946614 
# eelgrass_algae_barebare/algae     -2.5598736    1.3522715
# eelgrass_algae_barebare           -1.0105410    0.3946949 

# 
# visibilityhigh                    -0.5452848    1.2946614
# visibilitylow                     -2.9637975    1.3877053
# visibilitymedium                  -1.4121614    1.0548769 

b_estimate_cover<-c(plogis(2.3462651),plogis(-0.5452848),plogis(-2.5598736),plogis(-1.0105410))

b_se_lower_cover<-c(plogis(2.3462651-2*1.1650095),
plogis(-0.5452848-2*1.2946614),
plogis(-2.5598736-2*1.3522715),
plogis(-1.0105410-2*0.3946949))

b_se_upper_cover<-c(plogis(2.3462651+2*1.1650095),
plogis(-0.5452848+2*1.2946614),
plogis(-2.5598736+2*1.3522715),
plogis(-1.0105410+2*0.3946949))


b_estimate_vis<-c(plogis(-0.5452848),plogis(-1.4121614),plogis(-2.9637975))

b_se_lower_vis<-c(plogis(-0.5452848-2*1.2946614),
plogis(-1.4121614-2*1.0548769),
plogis(-2.9637975-2*1.3877053))

b_se_upper_vis<-c(plogis(-0.5452848+2*1.2946614),
plogis(-1.4121614+2*1.0548769),
plogis(-2.9637975+2*1.3877053))

# convert out of logit space
b_estimate<-c(plogis(-2.2085646),plogis(0.3460405),plogis(-1.7925503),plogis(-0.5452848),plogis(-1.6230180))

b_se_lower<-c(plogis(-2.2085646-2*1.418819),
plogis(0.3460405-2*1.143079),
plogis(-1.7925503-2*1.247346),
plogis(-0.5452848-2*1.294661),
plogis(-1.6230180-2*1.371282))

b_se_upper<-c(plogis(-2.2085646+2*1.418819),
plogis(0.3460405+2*1.143079),
plogis(-1.7925503+2*1.247346),
plogis(-0.5452848+2*1.294661),
plogis(-1.6230180+2*1.371282))

b_habitat<-c('eelgrass','mud flat','oyster flipbag','clam nets','oyster on bottom')
b_farm<-c('unfarmed habitat','unfarmed habitat','farm habitat','farm habitat','farm habitat')
benthic_mod_plot<-data.frame(b_estimate,b_habitat,b_farm)


benthic_mod_plot$b_habitat <- factor(benthic_mod_plot$b_habitat, 
                         levels=c("eelgrass",
                                  "mud flat",
                                  "oyster flipbag",
                                  "clam nets",
                                  "oyster on bottom"))

benthic_mod_plot$b_farm <- factor(benthic_mod_plot$b_farm, 
                         levels=c("unfarmed habitat",
                                  "farm habitat"))
# make figure 4
p1<-ggplot(data = benthic_mod_plot,aes(x=b_habitat,y=b_estimate,colour=b_farm))+
  geom_point(size=4)+
  scale_color_brewer(palette = "Set2")+
  theme_classic()+ 
  labs(x="habitat type",
       y="feeding probability",
       title = "crabs")+
  scale_y_continuous(breaks = c(0,0.25,0.5,0.75,1.0),
                     labels = c(0,0.25,0.5,0.75,1.0),
                     expand = c(0, 0),
                     limits = c(0,1))+
  guides(color=guide_legend(title="habitat category"))+
  geom_errorbar(aes(ymin=b_se_lower,ymax=b_se_upper,color=b_farm,width=0.2))

ggsave(here("figures", "fig_05_benthic_model_habitat_type.png"),width = 7,height = 7,units = "in")
```


model coefficients plots demersal
```{r model coefficients plots demersal}

demersal_model_top <- glm(feeding_behavior ~ 0 + habitat_type, data = demersal, family = binomial)

summary(demersal_model_top)$coefficients


# demersal  
#                               Estimate Std. Error
# habitat_typeeelgrass        -2.2823824  0.4694765
# habitat_typesediment        -0.6931472  0.2970443
# habitat_typeFlipbag         -0.7282385  0.2301210
# habitat_typeClam            -2.3272777  0.5238227
# habitat_typeOyster OnBottom -2.3025851  0.6055300

d_habitat<-c('eelgrass','mud flat','oyster flipbag','clam nets','oyster on bottom')

# convert out of logit space
d_estimate<-c(plogis(-2.2823824),plogis(-0.6931472),plogis(-0.7282385),plogis(-2.3272777),plogis(-2.3025851))

d_se_lower<-c(plogis(-2.2823824-2*0.4694765),
plogis(-0.6931472-2*0.2970443),
plogis(-0.7282385-2*0.2301210),
plogis(-2.3272777-2*0.5238227),
plogis(-2.3025851-2*0.6055300))

d_se_upper<-c(plogis(-2.2823824+2*0.4694765),
plogis(-0.6931472+2*0.2970443),
plogis(-0.7282385+2*0.2301210),
plogis(-2.3272777+2*0.5238227),
plogis(-2.3025851+2*0.6055300))

d_habitat<-c('eelgrass','mud flat','oyster flipbag','clam nets','oyster on bottom')
d_farm<-c('unfarmed habitat','unfarmed habitat','farm habitat','farm habitat','farm habitat')
demersal_mod_plot<-data.frame(d_estimate,d_habitat,d_farm)


demersal_mod_plot$d_habitat <- factor(demersal_mod_plot$d_habitat, 
                         levels=c("eelgrass",
                                  "mud flat",
                                  "oyster flipbag",
                                  "clam nets",
                                  "oyster on bottom"))

demersal_mod_plot$d_farm <- factor(demersal_mod_plot$d_farm, 
                         levels=c("unfarmed habitat",
                                  "farm habitat"))
# make figure 3
p2<-ggplot(data = demersal_mod_plot,aes(x=d_habitat,y=d_estimate,colour=d_farm))+
  geom_point(size=4)+
  scale_color_brewer(palette = "Set2")+
  theme_classic()+ 
  labs(x="habitat type",
       y="feeding probability",
       title = "demersal fish")+
    scale_y_continuous(breaks = c(0,0.25,0.5,0.75,1.0),
                     labels = c(0,0.25,0.5,0.75,1.0),
                     expand = c(0, 0),
                     limits = c(0,1))+
  guides(color=guide_legend(title="habitat category"))+
  geom_errorbar(aes(ymin=d_se_lower,ymax=d_se_upper,color=d_farm,width=0.2))

ggsave(here("figures", "fig_04_demersal_model_habitat_type.png"),width = 7,height = 7,units = "in")
```

model coefficients plots pelagic
```{r model coefficients plots pelagic}

# switch order of fixed effects to get the correct coefficient outputs for each fixed effect. 
pelagic_model_top<- glmer(feeding_behavior ~ 0 + habitat_type + eelgrass_algae_bare + (1 | Site), data = pelagic, family = binomial)

summary(pelagic_model_top)$coefficients

# pelagic
#                                 Estimate Std. Error 
# habitat_typeeelgrass          -0.58956672  0.6126456
# habitat_typesediment          -0.79780051  0.4746709 
# habitat_typeFlipbag           -0.38301527  0.4008872 
# habitat_typeClam              -0.05703813  0.4962638 
# habitat_typeOyster OnBottom   -0.69372562  0.6052590 
# 
# eelgrass_algae_barealgae      -0.05696037  0.4962427 
# eelgrass_algae_barebare       -0.72450949  0.4120620 
# eelgrass_algae_barebare/algae -1.42723666  0.6053752 
# eelgrass_algae_bareeelgrass   -0.40891847  0.5659252 

p_estimate_cover<-c(plogis(-0.05696037),plogis(-0.72450949),plogis(-1.42723666),plogis(-0.40891847))

p_se_lower_cover<-c(plogis(-0.05696037-2*0.4962427),
plogis(-0.72450949-2*0.4120620),
plogis(-1.42723666-2*0.6053752),
plogis(-0.40891847-2*0.5659252))

p_se_upper_cover<-c(plogis(-0.05696037+2*0.4962427),
plogis(-0.72450949+2*0.4120620),
plogis(-1.42723666+2*0.6053752),
plogis(-0.40891847+2*0.5659252))





# convert out of logit space
p_estimate<-c(plogis(-0.58956672),plogis(-0.79780051),plogis(-0.38301527),plogis(-0.05703813),plogis(-0.69372562))

p_se_lower<-c(plogis(-0.58956672-2*0.6126456),
plogis(-0.79780051-2*0.4746709),
plogis(-0.38301527-2*0.4008872),
plogis(-0.05703813-2*0.4962638),
plogis(-0.69372562-2*0.6052590))

p_se_upper<-c(plogis(-0.58956672+2*0.6126456),
plogis(-0.79780051+2*0.4746709),
plogis(-0.38301527+2*0.4008872),
plogis(-0.05703813+2*0.4962638),
plogis(-0.69372562+2*0.6052590))


p_habitat<-c('eelgrass', 'mud flat', 'oyster flipbag', 'clam nets' , 'oyster on bottom')
p_farm<-c('unfarmed habitat','unfarmed habitat','farm habitat','farm habitat','farm habitat')
pelagic_mod_plot<-data.frame(p_estimate,p_habitat,p_farm)


pelagic_mod_plot$p_habitat <- factor(pelagic_mod_plot$p_habitat, 
                         levels=c("eelgrass",
                                  "mud flat",
                                  "oyster flipbag",
                                  "clam nets",
                                  "oyster on bottom"))

pelagic_mod_plot$p_farm <- factor(pelagic_mod_plot$p_farm, 
                         levels=c("unfarmed habitat",
                                  "farm habitat"))
# make figure 2
ggplot(data = pelagic_mod_plot,aes(x=p_habitat,y=p_estimate,colour=p_farm))+
  geom_point(size=4)+
  scale_color_brewer(palette = "Set2")+
  theme_classic()+ 
  labs(x="habitat type",
       y="feeding probability",
       title = "pelagic fish")+
    scale_y_continuous(breaks = c(0,0.25,0.5,0.75,1.0),
                     labels = c(0,0.25,0.5,0.75,1.0),
                     expand = c(0, 0),
                     limits = c(0,1))+
  guides(color=guide_legend(title="habitat category"))+
  geom_errorbar(aes(ymin=p_se_lower,ymax=p_se_upper,color=p_farm,width=0.2))

ggsave(here( "figures", "fig_02_pelagic_model_habitat_type.png"),width = 7,height = 7,units = "in")
```


pelagic model cover type plot
```{r pelagic model cover type plot}


pelagic_model_top<- glmer(feeding_behavior ~ 0 + eelgrass_algae_bare + habitat_type + (1 | Site), data = pelagic, family = binomial)

summary(pelagic_model_top)$coefficients

# pelagic
# 
# eelgrass_algae_barebare       -0.72450949  0.4120620 
# eelgrass_algae_barebare/algae -1.42723666  0.6053752 
# eelgrass_algae_barealgae      -0.05696037  0.4962427 
# eelgrass_algae_bareeelgrass   -0.40891847  0.5659252 



# convert out of logit space
p_estimate_cover<-c(plogis(-0.72450949),plogis(-1.42723666),plogis(-0.05696037),plogis(-0.40891847))

p_se_lower_cover<-c(plogis(-0.72450949-2*0.4120620),
plogis(-1.42723666-2*0.6053752),
plogis(-0.05696037-2*0.4962427),
plogis(-0.40891847-2*0.5659252))

p_se_upper_cover<-c(plogis(-0.72450949+2*0.4120620),
plogis(-1.42723666+2*0.6053752),
plogis(-0.05696037+2*0.4962427),
plogis(-0.40891847+2*0.5659252))



p_cover_data<-c('bare', 'bare/algae', 'algae', 'eelgrass')
pelagic_mod_plot<-data.frame(p_cover_data,p_estimate_cover)


pelagic_mod_plot$p_cover_data <- factor(pelagic_mod_plot$p_cover_data, 
                         levels=c("bare",
                                  "bare/algae",
                                  "algae",
                                  "eelgrass"))



# make figure 5
ggplot(data = pelagic_mod_plot,aes(x=p_cover_data,y=p_estimate_cover))+
  geom_point(size=4)+
  theme_classic()+ 
  labs(x="cover type",
       y="feeding probability",
       title = "pelagic fish")+
    scale_y_continuous(breaks = c(0,0.25,0.5,0.75,1.0),
                     labels = c(0,0.25,0.5,0.75,1.0),
                     expand = c(0, 0),
                     limits = c(0,1))+
 theme(legend.position = "none")+
  geom_errorbar(aes(ymin=p_se_lower_cover,ymax=p_se_upper_cover,width=0.2))

ggsave(here( "figures", "fig_03_pelagic_model_cover_type.png"),width = 7,height = 7,units = "in")
```





benthic model cover type plot
```{r benthic model cover type plot}


top_model_benthic <- glm(feeding_behavior ~ 0 + habitat_type + visibility + eelgrass_algae_bare, data = benthic, family = binomial)

summary(top_model_benthic)$coefficients

# benthic
# 
# eelgrass_algae_barebare           -1.0105410    0.3946949 
# eelgrass_algae_barebare/algae     -2.5598736    1.3522715
# eelgrass_algae_barealgae          -0.5452848    1.2946614 
# eelgrass_algae_bareeelgrass        2.3462651    1.1650095 

# convert out of logit space
p_estimate_cover<-c(plogis(-1.0105410),plogis(-2.5598736),plogis(-0.5452848),plogis(2.3462651))

p_se_lower_cover<-c(plogis(-1.0105410-2*0.3946949),
plogis(-2.5598736-2*1.3522715),
plogis(-0.5452848-2*1.2946614),
plogis(2.3462651-2*1.1650095))

p_se_upper_cover<-c(plogis(-1.0105410+2*0.3946949),
plogis(-2.5598736+2*1.3522715),
plogis(-0.5452848+2*1.2946614),
plogis(2.3462651+2*1.1650095))



p_cover_data<-c('bare', 'bare/algae', 'algae', 'eelgrass')
pelagic_mod_plot<-data.frame(p_cover_data,p_estimate_cover)


pelagic_mod_plot$p_cover_data <- factor(pelagic_mod_plot$p_cover_data, 
                         levels=c("bare",
                                  "bare/algae",
                                  "algae",
                                  "eelgrass"))



# make figure 6
ggplot(data = pelagic_mod_plot,aes(x=p_cover_data,y=p_estimate_cover))+
  geom_point(size=4)+
  theme_classic()+ 
  labs(x="cover type",
       y="feeding probability",
       title = "crabs")+
    scale_y_continuous(breaks = c(0,0.25,0.5,0.75,1.0),
                     labels = c(0,0.25,0.5,0.75,1.0),
                     expand = c(0, 0),
                     limits = c(0,1))+
 theme(legend.position = "none")+
  geom_errorbar(aes(ymin=p_se_lower_cover,ymax=p_se_upper_cover,width=0.2))

ggsave(here( "figures", "fig_06_benthic_model_cover_type.png"),width = 7,height = 7,units = "in")
```
